---
title: "RMS Titanic"
author: "(Your name here)"
date: 2020-
output:
  github_document:
    toc: true
prerequisites:
  - e-vis01-bar-charts
---

*Purpose*: Most datasets have at least a few variables. Part of our task in analyzing a dataset is to understand trends as they vary across these different variables. Unless we're careful and thorough, we can easily miss these patterns. In this challenge you'll analyze a dataset with a small number of categorical variables and try to find differences among the groups.

*Reading*: (Optional) [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic.

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission
<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**


```{r setup}
library(tidyverse)

df_titanic <- as_tibble(Titanic)
```

*Background*: The RMS Titanic sank on its maiden voyage in 1912; about 67% of its passengers died.

# First Look
<!-- -------------------------------------------------- -->

### __q1__ Perform a glimpse of `df_titanic`. What variables are in this dataset?

```{r q1-task}
## TASK: Perform a `glimpse` of df_titanic
glimpse(df_titanic)
```

**Observations**:
There is 32 rows and 5 columns in the data set of df_titanic. The 5 variables are Class, Sex, Age, Survived and n. Class variable show the passenger's class belong to 1st, 2nd, 3rd or crew. Sex variable show the passenger is Male or Female. Age variable show the passenger is child or adult. Survived variable show passenger survived or not. And the last variable n show number of people in each specific situaition of class, sex age and suvival status. In this data set, each row represent a group of people. 


### __q2__ Skim the [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic, and look for a total count of souls aboard. Compare against the total computed below. Are there any differences? Are those differences large or small? What might account for those differences?

```{r q2-summary}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_titanic %>% summarize(total = sum(n))
```

**Observations**:
The information of Wikipedia show that the total number of passengers and crews aboard was 2208. However, the database records only 2201 people, which mean 7 are missing. The reason for it would be the error during the data collection or copying process.


### __q3__ Create a plot showing the count of persons who *did* survive, along with aesthetics for `Class` and `Sex`. Document your observations below.

*Note*: There are many ways to do this.

```{r q3-task}
## TASK: Visualize counts against `Class` and `Sex`
df_titanic %>%
  filter(Survived == "Yes")%>%
  ggplot(aes (x = Class, y = n, fill = Sex)) +
  geom_col(position = "dodge") 
```

**Observations**:
The male crew has the largest number of survivors with about 190 people.The second largest survival group is 1st class female with about 140 people. The lowest survival group is 2nd class males, followed by female crews. The number of female and male survivors is almost the same in 3rd class. 
The overall data show that there are more female survivors except the crew group. The number of male crew members who survived is more than all the other class combined. However it's hard to draw conclusion because it didn't show the total number of passenger abroad in each group. 

# Deeper Look
<!-- -------------------------------------------------- -->

Raw counts give us a sense of totals, but they are not as useful for understanding differences between groups. This is because the differences we see in counts could be due to either the relative size of the group OR differences in outcomes for those groups. To make comparisons between groups, we should also consider *proportions*.[1]

The following code computes proportions within each `Class, Sex, Age` group.

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop
```

### __q4__ Replicate your visual from q3, but display `Prop` in place of `n`. Document your observations, and note any new/different observations you make in comparison with q3. Is there anything *fishy* in your plot?

```{r q4-task}
df_prop %>%
  filter(Survived == "Yes")%>%
  ggplot(aes (x = Class, y = Prop, fill = Sex)) +
  geom_col(position = "dodge") 

```
```{r}
df_prop %>%
  filter(Survived == "Yes")%>%
  ggplot(aes (x = Class, y = Prop, fill = Sex)) +
  geom_col() 
```


**Observations**:

- The proportional of male and female that survive is 1, which is different with the proportion number in the datasheet above. The group of 3rd and crew seem better.
- There is anything *fishy* going on in the plot the Prop of the bar represent the sum of proportion in all age of the class, rather than the individual proportion in the data set. Without the "position = dodge", The bar show that the proportion number is over 3 for the 1st class. 

### __q5__ Create a plot showing the group-proportion of occupants who *did* survive, along with aesthetics for `Class`, `Sex`, *and* `Age`. Document your observations below.

*Hint*: Don't forget that you can use `facet_grid` to help consider additional variables!

```{r q5-task}
df_prop %>%
  filter(Survived == "Yes")%>%
  ggplot(aes (x = Class, y = Prop, fill = Sex)) +
  geom_col(position = "dodge") +
  facet_wrap(~Age)

```

**Observations**:
- The proportion of female survivors is much higher than male survivors, especially among 1st-class passengers, followed by 2nd-class females and female crew members.
For children, all children in 1st and 2nd class survived, while 3rd-class children had a much lower survival proportion. A similar pattern can be seen among adults: 1st-class males and females had higher survival proportions than those in the other classes. Among adult males, the 2nd-class group had a particularly low survival proportion, while 3rd-class males had a slightly higher proportion and followed by the male crew. Overall, age, gender, and class all appear to affect the proportion of survivors in different level. There might be other factor that could affect it that didn't show in the dataset.
- The comparison of two plot show the problem is exactly in the combination of age group. All the number and bar look better and accurate after the separation of adult and child. 

# Notes
<!-- -------------------------------------------------- -->

[1] This is basically the same idea as [Dimensional Analysis](https://en.wikipedia.org/wiki/Dimensional_analysis); computing proportions is akin to non-dimensionalizing a quantity.
